Extracellular potentials generated by axonal projections are shaped by patterns of bifurcations and terminations

### 2016-02-01 doc/manuscript.tex
some small changes from RK

EFPs in the brain were long thought to be primarily synaptic in origin
(Buzsáki et al., 2012). As a consequence, many modeling studies focus on
the extracellular fields induced by synaptic currents on the dendrites
and soma of the postsynaptic neuron (Holt and Koch, 1999; Gold et al.,
2006; Lindén et al., 2010, 2011; Einevoll et al., 2013; Fernández-Ruiz
et al., 2013). However, a number of recent data analysis and modeling
efforts have revealed that active, non-synaptic membrane currents can
play an important role in generating EFPs (Ray and Maunsell, 2011;
Belluscio et al., 2012; Schomburg et al., 2012; Reimann et al., 2013;
Anastassiou et al., 2015), including far reaching potentials detectable
at the scalp (Telenczuk et al., 2011).

One of the reasons for the traditional assumption that axonal currents
contribute little to the EFP is that the far field of an action
potential traveling along an idealized straight axon is quadrupolar,
meaning that its far field decays faster with distance than synaptic
sources which are thought to be dipolar (Nunez and Srinivasan, 2006). In
experimental studies that recoded the EFP of axonal responses, the EFP
often has a dipolar structure. For example, Blot and Barbour (2014)
report an EFP with the characteristic dipolar structure recorded from
the vicinity of cerebellar Purkinje cell axons.

@@

The aim of this study is to show how these effects in the EFP of axons
are influenced by the axons' anatomical structure. In particular, we aim
to explain how typical projection patterns in which axons bifurcate and
then terminate in their projection area affect the EFP. Such axon
bundles, sometimes called nerves or facicles, exist throughout the
peripheral and central nervous system of most animals (Nornes and Das,
1972; Goodman et al., 1984; Hentschel and Ooyen, 1999; Kandel et al.,
2000). The white matter of the mammalian brain can be viewed as an
agglomeration of such bundles (Schüz and Braitenberg, 2002).

\section{Results}\label{results}

We characterize three principal effects of axon bundle structure on the
EFP. Firstly, the low-frequency components of the EFP are governed by
the densities of bifurcations and terminations and can have a dipolar
structure. Secondly, the high-frequency components are governed by the
local fiber density. Furthermore we show that the low-frequency
components exceed the high-frequency components in spatial reach. In
particular, the dipolar low-frequency components exceed the reach of the
presumed quadrupolar nature of axonal EFPs.

These properties are first investigated by numerical model based on a
forward simulation of the extracellular field potential (Holt and Koch,
1999; Gold et al., 2006). This first model includes a large-scale
multicompartment model of the axon population. We then show a concrete
example of the effects in experimental data using as a set of multi-site
in-vivo electrophysiological recordings from the barn owl brain stem.
Finally, we show the mechanisms underlying the observed effects by means
of a second, analytically tractable, model of a generic axon bundle.

\emph{Axonal projections generate a dipole-like field potential.} To
understand the effect of axon geometry on the EFP, we simulated a fiber
bundle that at first runs at constant diameter and without bifurcations,
then reaches a bifurcation zone with an increasing number of fibers,
before finally terminating (Figure 1A and 1D). The bundle fibers were
stimulated with a background spontaneous rate, upon which a pulse of
increased activity was added. The resulting extracellular potential was
calculated for several locations along the axon trunk. The spatial
behaviour of the low-frequency component of the EFP differed strongly
from that of the high-frequency component, so we analyzed them
separately. Low-frequency responses are shown in the top half of Figure
1, and low-frequency responses are shown in the lower half.

@@

\hyperdef{}{ref-Buzsaki2012Origin}{}
Buzsáki G, Anastassiou CA, Koch C (2012) The origin of extracellular
fields and currents --- EEG, ECoG, LFP and spikes. Nature Reviews
Neuroscience 13:407--420 Available at:
\url{http://dx.doi.org/10.1038/nrn3241}.

@@

\hyperdef{}{ref-Einevoll2013Modelling}{}
Einevoll GT, Kayser C, Logothetis NK, Panzeri S (2013) Modelling and
analysis of local field potentials for studying the function of cortical
circuits. Nature Reviews Neuroscience 14:770--785 Available at:
\url{http://dx.doi.org/10.1038/nrn3599}.

@@

\hyperdef{}{ref-Telenczuk2011Highfrequency}{}
Telenczuk B, Baker SN, Herz AVM, Curio G (2011) High-frequency EEG
covaries with spike burst patterns detected in cortical neurons. Journal
of Neurophysiology 105:2951--2959 Available at:
\url{http://dx.doi.org/10.1152/jn.00327.2010}.

>>>>

EFPs in the brain were long thought to be primarily synaptic in origin
(Buzsáki et al., 2012). As a consequence, many modeling studies focus on
the extracellular fields induced by synaptic currents on the dendrites
and soma of the postsynaptic neuron (Holt and Koch, 1999; Gold et al.,
2006; Lindén et al., 2010, 2011; Einevoll et al., 2013; Fernández-Ruiz
et al., 2013). However, a number of recent data analysis and modeling
efforts have revealed that active, non-synaptic membrane currents can
play an important role in generating EFPs (Ray and Maunsell, 2011;
Belluscio et al., 2012; Schomburg et al., 2012; Reimann et al., 2013;
Anastassiou et al., 2015), including far reaching potentials detectable
at the scalp (Teleńczuk et al., 2011, 2015).

One of the reasons for the traditional assumption that axonal currents
contribute little to the EFP is that the far field of an action
potential traveling along an idealized straight axon is quadrupolar,
meaning that its far field decays faster with distance than synaptic
sources which are thought to be dipolar (Nunez and Srinivasan, 2006).
However, in experimental studies that recorded the EFP of axonal
responses, the EFP often has a dipolar structure. For example, Blot and
Barbour (2014) report an EFP with the characteristic dipolar structure
recorded from the vicinity of cerebellar Purkinje cell axons.

@@

This discrepancy raises the open question of how axons are able to
generate dipolar field potentials in the actual organism, when
conventional simulation results lead one to expect quardupolar
structures.

The aim of this study is to show how these effects in the EFP of axons
can be explained by the axons' anatomical structure. In particular, we
aim to explain how typical projection patterns in which axons bifurcate
and then terminate in their projection area lead to a dipolar EFP
structure. Such axon bundles, sometimes called nerves or facicles, exist
throughout the peripheral and central nervous system of most animals
(Nornes and Das, 1972; Goodman et al., 1984; Hentschel and Ooyen, 1999;
Kandel et al., 2000). The white matter of the mammalian brain can be
viewed as an agglomeration of such bundles (Schüz and Braitenberg,
2002).

These properties are first investigated by numerical model based on a
forward simulation of the extracellular field potential (Holt and Koch,
1999; Gold et al., 2006). This first model includes a large-scale
multicompartment model of the axon population. We then show a concrete
example of the effects in experimental data using as a set of multi-site
in-vivo electrophysiological recordings from the barn owl brain stem.
Finally, we show the mechanisms underlying the observed effects by means
of a second, analytically tractable, model of a generic axon bundle.
Results ==============

\emph{Axonal projections generate a dipole-like field potential.} To
understand how axons can generate a dipolar EFP, we simulated a parallel
fiber bundle that at first runs at constant diameter and without
bifurcations, then reaches a bifurcation zone with an increasing number
of fibers, and finally reaches a termination zone, in which the number
of fibers decreases to zero. (Figure 1A and 1D). The bundle fibers were
stimulated with a background spontaneous rate, upon which a pulse of
increased activity was added. To characterize the spatiotemporal
dynamics of the evoked EFP, the potential was calculated for several
locations along the axon trunk.

The spatial behaviour of the low-frequency component of the EFP differed
strongly from that of the high-frequency component, so we analyzed them
separately. Low-frequency responses are shown in the top half of Figure
1, and high-frequency responses are shown in the lower half.

@@

\hyperdef{}{ref-Buzsaki2012Origin}{}
Buzsáki G, Anastassiou CA, Koch C (2012) The origin of extracellular
fields and currents--EEG, ECoG, LFP and spikes. Nature reviews
Neuroscience 13:407--420 Available at:
\url{http://dx.doi.org/10.1038/nrn3241}.

@@

\hyperdef{}{ref-Einevoll2013Modelling}{}
Einevoll GT, Kayser C, Logothetis NK, Panzeri S (2013) Modelling and
analysis of local field potentials for studying the function of cortical
circuits. Nature reviews Neuroscience 14:770--785 Available at:
\url{http://dx.doi.org/10.1038/nrn3599}.

@@

\hyperdef{}{ref-Telenczuk2011Highfrequency}{}
Teleńczuk B, Baker SN, Herz AVM, Curio G (2011) High-frequency EEG
covaries with spike burst patterns detected in cortical neurons. Journal
of Neurophysiology 105:2951--2959 Available at:
\url{http://dx.doi.org/10.1152/jn.00327.2010}.

\hyperdef{}{ref-Telenczuk2015Correlates}{}
Teleńczuk B, Baker SN, Kempter R, Curio G (2015) Correlates of a single
cortical action potential in the epidural EEG. NeuroImage 109:357--367
Available at: \url{http://dx.doi.org/10.1016/j.neuroimage.2014.12.057}.
